---
title: '**Federmesser vs. Bromme**'
author:
- Dr. Christian Steven Hoggard (University of Southampton)
- Prof. Felix Riede (Aarhus University)
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
subtitle: A morphometric and technological analysis of Final Palaeolithic material
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="cairo_pdf")
```

```{r chunk1, echo = FALSE, message = FALSE}
library(rio)
library(tidyverse)
library(gt)
library(FactoMineR)
library(factoextra)
```

# About the data

This document is an exploratory analysis of blade, core, tanged point and backed point data from Denmark, Sweden and Germany collected from 2017-2019. All data is stored on GitHub in an open-access repository: https://github.com/CSHoggard/-tech-data. The data is brought into this RMarkdown document via the `rio::import` package. The `tidyverse`, `factoMineR`, `factoextra` and `gt` packages were also used throughout this document; please ensure both are downloaded to explore the data in RStudio.

In total there are 599 complete artefacts. See Table 1 and Table 2 for more information.

```{r chunk2, echo = FALSE, message = FALSE}

data <- import("https://raw.githubusercontent.com/CSHoggard/-tech-data/master/Hoggard_and_Riede_2020.csv", na.strings=c(""), setclass = "tbl_df", encoding = "UTF-8")

cols <- c("CODE", 
          "CONTEXT", 
          "COUNTRY", 
          "CONDITION", 
          "ARTEFACT", 
          "SUB_CATEGORY", 
          "RECOVERY_METHOD", 
          "CLASSIFICATION", 
          "BP_ASSOCIATION", 
          "TP_ASSOCIATION", 
          "ABS_DATE_METHOD", 
          "RELAT_DATE_METHOD", 
          "RELAT_DATE_CHRONO", 
          "RAW_MAT", 
          "DORS_BLADE_PROF", 
          "BLADE_DET", 
          "BLADE_CURV", 
          "DORSAL_PATTERN", 
          "BULB_MORPH", 
          "CONUS_FORM", 
          "BUTT_MORPH", 
          "BUTT_PREP_1", 
          "BUTT_PREP_2", 
          "CORE_MORPH", 
          "PLAT_REJUV", 
          "CORE_METHOD", 
          "CORE_DIRECTIONALITY", 
          "CORE_TABLET_REJUV", 
          "CORE_FLAKE_REJUV", 
          "CORE_FRONT_REJUV", 
          "CORE_DIST_REJUV", 
          "CORE_SIDE_REJUV", 
          "BURINATION")

data[cols] <- lapply(data[cols], factor)

data %>% 
  rename(Context = CONTEXT) %>%
  group_by(ARTEFACT) %>%
  count(Context) %>%
  pivot_wider(names_from = ARTEFACT, values_from = n) %>%
  gt() %>%
  tab_header(title = "Table 1. Artefact vs. Context (n = 599)")
```

Sassenholz makes up the greatest proportion of the database featuring 219 artefacts, followed by the Rietberg complex (111), Søvind (93) and Häcklingen (67).

```{r chunk3, echo = FALSE, message = FALSE}
data %>% 
  rename(Classification = CLASSIFICATION) %>%
  group_by(ARTEFACT) %>%
  count(Classification) %>%
  pivot_wider(names_from = ARTEFACT, values_from = n) %>%
  gt() %>%
  tab_header(title = "Table 2. Artefact vs. Classification (n = 599)")
```

In total 383 originate from a presupposed Bromme context, representing 63.83% of all material. The Federmesser is here represented by 217 artefacts, representing 36.17% of all material.

```{r chunk4, echo = FALSE, message = FALSE}
blade_data <- filter(data, ARTEFACT == "Blade")
core_data <- filter(data, ARTEFACT == "Core")
bp_data <- filter(data, ARTEFACT == "Backed Point")
tp_data <- filter(data, ARTEFACT == "Tanged Point")
```

**Aims**  
* To examine the robustness of Federmesser and Bromme division as grounded on technological and morphometric characteristics.  
* To assess typologically-awkward sites e.g. Häcklingen.  

**Methodology**
The following methodology is employed here:  
* Technological characteristics are examined through visual summaries of data and Mann-Whitney pairwise tests (Bonferroni-correct p values).  
* Gini–Simpson and Morisita–Horn diversity indices are employed to examine intra-site variability (calculated using the function H in the vegetarian
package); past applications in Crema (2014) and Maiorano et al. (2020). 
* Multiple Correspondence Analysis (MCA) is then performed for all categorical technological data for each data type (core, blade, backed point, tanged point); this will examine individual contexts and their classification.   
* Principal Component Analysis (PCA) is performed for all quantitative (morphometric) data, again for each data type; this will also examine individual contexts and their classification.    
* Multiple Factor Analysis (MFA) will then be employed to test all variables in one analytical protocol (qualitative and quantitative).  
* A MANOVA is employed in this latter process for a statistical framework.